The Economics of Shotgun Marriage

### 2020-10-02 slides-march-20/slides-old.tex
Current stage

\usepackage{fixmath}
\usepackage{tikz}
\usetikzlibrary{shapes,arrows}
\usetikzlibrary{positioning}
\tikzstyle{block} = [rectangle, draw, rounded corners]
\tikzstyle{line} = [draw, -latex']

@@

% There is a huge social issue of single mothers
% You typically associate it with unplanned teenage pregnancies of low-income and low-education people
% But looking at just single mothers or young couples is an understatement
% Unplanned pregnancy expresses itself even for married people: not all marriages are created equal
% In later ages unplanned pregnancies force couple to marry
% 
% Even for college 
% In more adult age this is pronounced by shotgun marriage
% Chetty stuff: share of single mothers is the best predictor of inequality and so on

@@

\begin{frame}
\frametitle{Unplanned Pregnancy $\Rightarrow$ Shotgun Marriage}
\begin{itemize}
\item Many people marry shortly after a baby appears
\item These marriages perform worse in the data 
\begin{itemize}
\item A lot worse for college grads, a bit worse for high school grads
\item $\sim10\%$ marry while being pregnant
\item Not just a teenage thing, peaks around 30
\end{itemize}
\item Understanding why helps to understand the value of marriage
\item Questions:
\begin{itemize}
\item What is the aggregate role of unplanned pregnancies?
\item What causes people to enter such marriages?
\item Why they perform worse?
\end{itemize}
\end{itemize}
\end{frame}

\begin{frame}[label=pic0]
\frametitle{Divorced By Timings}
\begin{center}
\includegraphics[width=0.85\linewidth]{share_div_march5.eps}
\end{center}
\end{frame}

\begin{frame}[label=pic0]
\frametitle{Hazards}
\begin{center}
\includegraphics[width=0.85\linewidth]{haz_kf_march5.eps}
\end{center}
\end{frame}

\begin{frame}[label=pic0]
\frametitle{Shares}
\begin{center}
\includegraphics[width=0.85\linewidth]{share_kj_march5.eps}
\end{center}
\end{frame}

\begin{frame}[plain]
\frametitle{What I Do}
\begin{itemize}
\item Quantify empirical difference of marriage performance by the timing of childbirth
\item Estimate a structural model matching this difference
\item Use the model to study:
\begin{itemize}
\item aggregate impact of unplanned childbirths 
\item importance of fertility for marriage
\item difference between education groups
\item policy counterfactuals
\end{itemize}
\end{itemize}
\end{frame}

\begin{frame}[plain]
\frametitle{Preview of Results}

\begin{enumerate}

\item Data: 
{\small
\begin{center}
\begin{tabular}{|c|c|c|c|}\hline
 & \multicolumn{2}{|c|}{\textit{Share of divorced among:}} & \\
& Marriage First & Kids First & Ratio \\\hline
College &  5.3\% &  14.7\% & \textbf{2.77} \\
High School &  13.9\% & 17.2\% &  \textbf{1.24}  \\\hline
\end{tabular}
\end{center}
}

\item Quantifying:
\begin{itemize}
\item Unplanned pregnancy changes outside options in family bargaining
\item Effect is present generically, its size identifies amount of distress
\item To estimate, assume optimal fertility timing after marriage
\end{itemize}

\item Exploring:
\begin{itemize}
\item Unplanned pregnancies account for $\sim50\%$ of early divorces
\item Marriage market (meeting probabilities) mostly drive the college -- high school difference
\end{itemize}

\end{enumerate}
\end{frame}

%
%\begin{frame}
%\frametitle{[Distribution of age at shotgun marriage in my def]}
%
%\end{frame}

%
%\begin{frame}
%\frametitle{Main Results}
%\begin{itemize}
%\item Empirical:
%\begin{itemize}
%\item Shotgun marriage $\Rightarrow$ divorced $\sim1.8$ more often
%\item Among college graduates: $\sim3$ times more often
%\item Differences are present after controlling for lots of observables
%\item Present is many data sets: ACS, SIPP, NSFG, NSFH
%\end{itemize}
%\item From the model:
%\begin{itemize}
%\item Disentangling two main mechanisms:
%\begin{itemize}
%\item Shotgun marriage couples are different in characteristics
%\item Unexpected pregnancy creates large unplanned costs
%\end{itemize}
%\item Expensive eraly-life childcare explains a lot
%\item Higher marginal costs of childcare for skilled women
%\item $\sim10\%$ of couples marry because of unplanned pregnancy
%\item These couples are the least stable
%\end{itemize}
%\item Missing pieces (TBD):
%\begin{itemize}
%\item Quantifying implications
%\item Quantifying changes over time
%\end{itemize}
%\end{itemize}
%\end{frame}

\begin{frame}
\frametitle{General Importance}
\begin{itemize}
\item Unintended pregnancies $\Rightarrow$ people to pick less favorable partners, ``inefficient'' marriages from ex-ante prospective
\item Child development:
\begin{itemize}
\item Single mothers or blended families $\Rightarrow$ inferior child's outcomes
\item Unstable marriages may be even worse but are often overlooked
\end{itemize}
\item Major impact on women, in particular on:
\begin{itemize}
\item Labor market: loss of skills
\item Divorces + children worsen re-marriage prospective
\item Total fertility and marriage rates
\end{itemize}
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{Literature}
Shotgun marriages are still a thing:
\begin{itemize}
\item Gibson-Davis et al (2016): 10\% midpregnancy marriages
%\item Akerlof et al (1996)
\end{itemize}
Lifecycle models:
\begin{enumerate}
\item Endogenous bargaining, marriage, divorce \& assets:
\begin{itemize}
\item Voena (2015) --- just divorce, Low et al (2018) --- both
\item Mazzocco (2007) --- seminal paper
\end{itemize}
%\begin{center}
%\textbf{Contribution:} endogenous + exogenous fertility
%\end{center}
\item Fertility:
\begin{itemize}
\item Planned: Sommer (2016)
\item Unplanned: Ejrnæs, Jørgensen (2018)
\end{itemize}
%\begin{center}
%\textbf{Contribution:} bargaining, marriage and divorce
%\end{center}
\item Child investment and marriage/divorce:
\begin{itemize}
\item Lafortune, Low (2018)
\end{itemize}
%\begin{center}
%\textbf{Contribution:} proper lifecycle, fertility
%\end{center}
\end{enumerate}
%
%\textit{Other relevant evidence:}
%\begin{itemize}
%\item Divorce rates by income: Boertien, Härkönen (2014)
%\item Child expenditures by income: Schneider et al (2014)
%\end{itemize}
%\item Computations: Judd et al (2014) (thank Luigi Bocola)

\begin{frame}
\frametitle{Updates Since November}
\begin{itemize}
\item Brand new codes in Python \& cuda
\item Remarriage of single mothers
\item Skills depreciation because of time to childcare
\item No unplanned pregnancies in marriage: restrictive but cleaner identification
\item Two separate estimations: college grads versus non-college grads
\item Decomposition exercises as a goal
\end{itemize}
\end{frame}

\section{Mechanism}

\begin{frame}
\frametitle{Framework}
\begin{itemize}
\item Bargaining model with outside option of being single. Together if:
\[V^{\text{fem},\text{married}} \geq V^{\text{fem},\text{single}} \ \& \ V^{\text{male},\text{married}} \geq V^{\text{male},\text{single}}\]
\item Make choices and have some margins of adjustment
\item Heterogeneous relationship quality (love shock), the best predictor of divorce
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{Unplanned Pregnancy Shock}
\begin{itemize}
\item Married couples can choose their fertility
\item However, sometimes a child arrives by a random shock $\Rightarrow$ distress
\item Size depends on how indifferent are people between different fertility timings
\item College people are more sensitive?
\item What about marriage/divorce?
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{Asymmetric Response to Shock}
People stay together if
\[V^{\text{fem},\text{married}} \geq V^{\text{fem},\text{single}} \ \& \ V^{\text{male},\text{married}} \geq V^{\text{male},\text{single}}\]
\vspace{-1em}
\begin{itemize}
\item Bargaining model: if stress is symmetric (all $V\downarrow$ the same way) neither more marriages nor more divorces happens
\item $\Rightarrow$ females lose more, males gain more
\item If marriages differ in quality, more low-quality marriages are created
\end{itemize}

Why do people enter low-quality marriages?
\begin{itemize}
\item Women:
\begin{itemize}
\item splitting of childcare costs
\item worsening of future marital prospectives
\end{itemize}
\item Men:
\begin{itemize}
\item loss of access to their children otherwise
\end{itemize}
\end{itemize}
\end{frame}

\section{Model}

\begin{frame}
\frametitle{Overview}
\begin{itemize}
\item Lifecycle model
\item Assets ($a$) and productivity shocks ($z^f$, $z^m$)
\item Labor supply for females with kids
\item Endogenous marriage and divorce (limited commitment)
\item Reduced form marriage market:
\[p^{\text{meet}}(\text{Age}) \text{ --- meeting a random partner with similar $a$ and $z$}\]
\item After marriage: endogenous fertility choice
\item Before marriage can have an unplanned child
\[p^{\text{unplanned}}(\text{Age}) \text{ --- arrival of unplanned kids}\]
\end{itemize}
\end{frame}

   \node [block, below = 1.5 cm of 2] (5) {\textbf{Kids First} \\[-0.5ex] \textbf{Couple $(ck)$}};
   \node [block, below = 0.8 cm of 3] (7) {No Kids \\[-0.5ex] Couple $(c)$};
   \node [block, below = 0.8 cm of 7] (8) {\textbf{Mariage First} \\[-0.5ex] \textbf{Couple  $(ck)$}};
   \draw[->] (1) to [bend left = 20] node [left] {\scriptsize Unplanned \\[-0.75ex] \scriptsize pregnancy} (2);
   \draw[->] (1) to [bend right = 20] node [right] {\scriptsize No \\[-0.75ex] \scriptsize pregnancy} (3);
   \draw[->] (2) to node [left] {\scriptsize Agree} node [right] {\scriptsize \textbf{Shotgun} \\[-0.75ex]  \scriptsize \textbf{marriage}} (5);
   \draw[->] (3) to node [right] {\scriptsize Agree} node [left]  {\scriptsize \textbf{Regular} \\[-0.75ex]  \scriptsize \textbf{marriage}} (7);
   \draw[->] (7) to node [right] {\scriptsize Kids by choice} (8);

    \node [block, above left   = 1.0 cm and -0.75 cm of 2] (4) {\,\, Single male $(m,s)$\,\,};
    \node [block, above left   = 1.6 cm and -0.75 cm of 2] (4-sm) {Single mother $(f,sk)$};
	\node [block, above right   = 1.0 cm and -0.75 cm of 3] (6) {\,\, Single male $(m,s)$\,\,};
    \node [block, above right   = 1.6 cm and -0.75 cm of 3] (6-f) {\,Single female $(f,s)$\,};   
    \draw[->] (2.180) to [bend left = 30] node [left] {\scriptsize Disagree} (4.270);
    \draw[->] (3.0) to [bend right = 30] node [right] {\scriptsize Disagree} (6.270);
     \draw[->] (6.180) to [bend left = 30] node [above] {\scriptsize Back}  (1.-12);
   \draw[->] (6-f.180) to [bend right = 30] node [above] {\scriptsize Back} (1.12);
   \draw[->,thick,dashed] (4-sm.90) to [bend left = 30] node [above] {\scriptsize Worse marriage prospectives}  (1.168);
   \draw[->] (4.0) to [bend right = 30] node [above] {\scriptsize Back} (1.192);

\begin{frame}
\frametitle{Modeling Shotgun Marriage}
\begin{center}
\vspace{-0.4cm}
\begin{tikzpicture}[every text node part/.style={align=center}, scale=0.75]
   % Place nodes
   \node [block] (1) {\large \textbf{People}  \\[-0.5ex]  \large  \textbf{Randomly Meet}};
   \node [block, below left = 1.5 cm and 0.0 cm of 1] (2) {\textit{Nash} \\[-0.5ex] \textit{Bargaining}};   
   \node [block, below right = 1.5 cm and 0.0 cm of 1] (3) {\textit{Nash} \\[-0.5ex] \textit{Bargaining}};

   \node [block, below = 1.5 cm of 2] (5) {\textbf{Kids First} \\[-0.5ex] \textbf{Couple} ($ck$)};
   \node [block, below = 0.8 cm of 3] (7) {No Kids \\[-0.5ex] Couple ($c$)};
   \node [block, below = 0.8 cm of 7] (8) {\textbf{Mariage First} \\[-0.5ex] \textbf{Couple} ($ck$)};
   \draw[->] (1) to [bend left = 20] node [left] {\scriptsize Unplanned \\[-0.75ex] \scriptsize pregnancy} (2);
   \draw[->] (1) to [bend right = 20] node [right] {\scriptsize No \\[-0.75ex] \scriptsize pregnancy} (3);
   \draw[->] (2) to node [left] {\scriptsize Agree} node [right] {\scriptsize \textbf{Shotgun} \\[-0.75ex]  \scriptsize \textbf{marriage}} (5);
   \draw[->] (3) to node [right] {\scriptsize Agree} node [left]  {\scriptsize \textbf{Regular} \\[-0.75ex]  \scriptsize \textbf{marriage}} (7);
   \draw[->] (7) to node [right] {\scriptsize Kids by choice} (8);   
    \node [block, above left   = 0.8 cm and -0.75 cm of 2] (4) {\,Single male \\[-0.5ex] ($m,s$)\,\,};
    \node [block, above left   = 1.8 cm and -0.75 cm of 2] (4-sm) {\,Single mom \\[-0.5ex]  ($f,sk$)};
	\node [block, above right   = 0.8 cm and -0.75 cm of 3] (6) {\,\,Single male\,\,\\[-0.5ex] ($m,s$)};
    \node [block, above right   = 1.8 cm and -0.75 cm of 3] (6-f) {Single female\\[-0.5ex] ($f,s$)\,};   
    \draw[->] (2.180) to [bend left = 30] node [left] {\scriptsize Disagree} (4.270);
    \draw[->] (3.0) to [bend right = 30] node [right] {\scriptsize Disagree} (6.270);
     \draw[->] (6.180) to [bend left = 30] node [above] {\scriptsize Back}  (1.-12);
   \draw[->] (6-f.180) to [bend right = 30] node [above] {\scriptsize Back} (1.12);
   \draw[->,thick,dashed] (4-sm.90) to [bend left = 30] node [above] {\scriptsize Worse marriage prospectives}  (1.168);
   \draw[->] (4.0) to [bend right = 30] node [above] {\scriptsize Back} (1.192);

\begin{frame}[label=utilities]
\frametitle{Utility Functions}
\begin{itemize}
\item Couple, $\theta$ --- Pareto weight, $\psi$ --- common love shock:
\[U = \theta^f \cdot u(c_f) + \theta^m \cdot u(c_m) + \psi,\]
\item Couple + child, $Q$ --- child quality (public good):
\[U = \theta^f \cdot {u}(c_f) + \theta^m \cdot {u}(c_m) + \psi + u^{\text{child}}(Q),\]
\item Child quality: $Q = F(x,1-l^f)$, $x$ --- money, $l^f$ --- female labor supply
\item Budget constraint ($C$ reflects returns to scale, $a$ --- assets):
 \[C(c_f,c_m) + a' + x = R\cdot a + l^f \cdot W^f_t(z_f) + W^m_t(z_m),\]
\end{itemize}
\hyperlink{value-functions}{\beamerbutton{Value Functions}}
\end{frame}

\begin{frame}
\frametitle{Children In The Model}
\begin{itemize}
\item People like having children (public good):
\[u^{\text{child}} = \phi + \alpha \cdot \frac{Q^{1-\sigma}}{1-\sigma}.\]
\item $Q$ is produced out of money and female time
\item $Q = \left[ x^{\lambda} + \kappa \cdot (1-l^f)^{\lambda}\right]^{\frac1\lambda}$
\item \textit{Subsistence constraint:} $Q \geq \bar{Q}$
\item $l^f\downarrow$ $\Rightarrow$ loss in current income + loss in productivity
\item Productivity process: $z^f_t = z^f_{t-1} + \varepsilon^f_{t}  - $ {\boldmath $f(l)$}.
\item In calibration $l \in \{0.25,0.5,0.75,0.9\}$
\item Corresponding to depreciation: $f(l) \in \{0.1,0.05,0.0,0.0\}$.
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{Marriage}
\begin{itemize}
\item People (f \& m) meet, get a draw of love shock $\psi$ and agree to marry if exists a Pareto weight $\theta$ such that
\[V^{f,\text{yes}}(\underset{\scriptsize +}{\theta\vphantom{\psi}},\underset{\scriptsize +}{\psi}) \geq V^{f,\text{no}}, \ \ \ \ V^{m,\text{yes}}(\underset{\scriptsize  -}{\theta\vphantom{\psi}},\underset{\scriptsize +}{\psi}) \geq V^{m,\text{no}}.\]
\item \textbf{Mechanism:} unplanned pregnancy 
\begin{enumerate}
\item[(1)] lowers current $V^{f,\text{no}}$: hard to be a single mother
\item[(2)] increases future $V^{m,\text{yes}}$ and $V^{f,\text{yes}}$: good to have children
\end{enumerate} 
$(1) + (2)$ $\Rightarrow$ the couple agrees to marry with lower $\psi$ and $\theta$.
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{Divorce}
\begin{itemize}
\item Shocks to $\psi$, $z^f$ and $z^m$
\item People cannot commit not to divorce
\item $\theta$ may be renegotiated, stays constant if not
\item Divorce happens if no $\theta$ exist such that
\[V^{f,\text{together}}(\theta,\psi) \geq V^{f,\text{divorce}}, \ \ \ \ V^{m,\text{together}}(\theta,\psi) \geq V^{m,\text{divorce}}.\]
\item Couples with lower $\psi$ or more unequal $z$ divorce more!
\end{itemize}
\end{frame}

%
%\begin{frame}
%\frametitle{Individual Values}
%
%\end{frame}

\begin{frame}
\frametitle{Discrete Decisions}
\begin{itemize}
\item Couple's problem defines $V^{c}_t$ (couple's value) and $(V^{f,c}_t,V^{m,c})$ (individual values).
\item In the last period $V^{c}_t = \theta^f u(c^f) + \theta^m u(c^m)$, $V^{i,c} = u(c^i)$.
\item Collective model (what I use): have a kid if 
\[ \theta^f u(c^{f,k}) + \theta^m u(c^{m,k})  + u^{k} \geq \theta^f u(c^{f,nk}) + \theta^m u(c^{m,nk}).\]
\item In fact this is ad hoc, and there is another version: have a kid if \textbf{both}
\[ u(c^{f,k}) + u^{k} \geq u(c^{f,nk})  \ \text{ and } \ u(c^{m,k}) + u^{k} \geq u(c^{m,nk}).\]
\item \textit{Decision making is not totally defined...}
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{Discrete Decisions + Rebargaining}
\begin{itemize}
\item Decision also depends on $\theta$
\item $k_\theta = \I(V^{c,k}_\theta \geq V^{c,nk}_\theta)$ (or the 2nd protocol)
\item Set of acceptable $\theta$ are
\[ \Theta_f =  \{ \theta : V^{f,ck}_\theta \cdot k_\theta  + V^{f,nk}_\theta \cdot  (1-k_\theta ) \geq V^{f,\text{divorce}} \}\]
\[ \Theta_m = \{ \theta : V^{m,ck}_\theta  \cdot  k_\theta+ V^{m,nk}_\theta  \cdot  (1-k_\theta ) \geq V^{m,\text{divorce}} \}\]
\[ \Theta_c = \Theta_f \cap \Theta_m.\]
\item Pick new $\theta$ if $\Theta_c$ is non-empty
\item But in this case $\Theta_c$ may even be disjoint
\item In this case use Marcet, Marimon (2019) idea and pick the closest $\theta$ to the original value
\item But things can become quirky...
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{Remarriage}
\begin{itemize}
\item If couple with a child divorces, she becomes a single mother
\item Single mothers still meet single males
\item If partners agree to marry they become a regular couple with a child
\item Males have a one-time utility penalty parameter for marrying a single mother. This affects bargaining:
\[NBS = \left[ V^{m,MK} - {\text{\boldmath$\phi_m$}} - V^{m,S}\right] \times  \left[ V^{f,MK} - V^{f,S}\right].\]
\item Rationalizes decrease in marital prospectives for females with children
\item Estimated by matching share of remarried women with vs without children
\end{itemize}
\end{frame}

\section{Fit and Results}
\begin{frame}
\frametitle{Strategy}
\begin{itemize}

\item Two calibrations: college and high school
\item I use Simulated Method of Moments to calibrate $\sim 11$ parameters:
\begin{itemize}
\item Preferences
\item Probabilities of transitions
\item Child quality technology
\end{itemize}
\item Income paths to match ACS (trend and variance)
\item The target moments are taken from ACS cross-section
\item Childcare expenditure share from CEX
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{College vs High School}
\begin{center}
\begin{tabular}{|l|c|c|c|}
& \multicolumn{2}{|c|}{Share of divorced, \%} & \\
& MF & KF & Ratio \\
Baseline: college & 5.4 & 14.0 &   2.6  \\
+ income process of HS & 3.9 & 15.3 & 4.0 \\
+ pregnancy probabilities of HS & 5.5 & 14.4 & 2.6 \\
\textbf{+ meeting probabilities of HS} & \textbf{8.9} & \textbf{12.6}  & \textbf{1.4} \\
+ love shock distribution of HS & 5.1 & 14.5 & 2.9 \\
+ single mother prospective of HS & 5.0 & 15.1 &  3.0 \\
Comparison: high school &  11.8 & 16.6 & 1.4\\
\end{tabular}
\end{center}
\end{frame}

\begin{frame}
\frametitle{Lower Child Investment Even without Selection}
\vspace{-2cm}
\begin{center}
\includegraphics[width=1.0\linewidth]{../main/x-time.pdf}
\end{center}
\end{frame}

\begin{frame}
\frametitle{Conclusions And Plans}
\begin{itemize}
\item What we observe about childcare costs generates the pattern
\item KF couples have lower $\psi$ and divorce more
\item Unplanned pregnancies account for $1/2$ of early-age divorces
\item Preparing a new iteration now:
\begin{itemize}
\item TBD: partition by income needs to be improved
\item TBD: decompose by different contributors ($z$, $\psi$, age)
\end{itemize}
\item Model allows to account for other stories:
\begin{itemize}
\item Labor market discrimination of unmarried mothers
\item Social stigma against unmarried cohabitation
\end{itemize}
\end{itemize}

@@

@@

\begin{frame}[label=pic0]
\frametitle{In One Picture (Old \& Extended Version)}
\begin{center}
\includegraphics[width=0.85\linewidth]{diff_dt_h.pdf}
\end{center}
\end{frame}

>>>>

\usepackage{tikz}
\usetikzlibrary{shapes,arrows}
\usetikzlibrary{positioning}
\tikzstyle{block} = [rectangle, draw, rounded corners]
\tikzstyle{line} = [draw, -latex']

@@

@@

\begin{frame}%[label=pic0]
\frametitle{In One Picture}
\begin{center}
\includegraphics[width=0.85\linewidth]{diff-by-dt.eps}
\end{center}
%\hyperlink{disc2}{\beamerbutton{dis}}
\end{frame}

\begin{frame}
\frametitle{Shotgun Marriages Perform Worse}
\begin{itemize}
\item Couples who have their kids before marriage divorce more:
\begin{itemize}
\item \textbf{Especially} among college graduates
\item Selection may not be the primary reason
\end{itemize}
\item This teaches us about heterogeneity of marriages:
\begin{itemize}
\item Raising children alone is hard
\item Unplanned pregnancies $\Rightarrow$ lower quality marriages
\item Still important despite access to contraception and abortion
\end{itemize}
\pause
\item \textit{Unexpected pregnancy often triggers marriage}
\item \textit{Unmarried unions with children are rare}
\begin{itemize}
\item \textit{2013: 3\% of children live with unmarried parents}
\item \textit{As opposed to 21\% with single mothers}
\end{itemize}
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{What I Do}

\begin{itemize}
\item Document the fact empirically. Lots of selection!
\item Construct and estimate a structural model 
\item Use it to understand observable differences in divorce rates
\end{itemize}

Main questions:
\begin{itemize}
\item What explains bad performance of shotgun marriage
\item Why college graduates are affected relatively more
\item Role of policies (childcare etc)
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{Main Results}
\begin{itemize}
\item Empirical:
\begin{itemize}
\item Shotgun marriage $\Rightarrow$ divorced $\sim1.8$ more often
\item Among college graduates: $\sim3$ times more often
\item Differences are present after controlling for lots of observables
\item Present is many data sets: ACS, SIPP, NSFG, NSFH
\end{itemize}
\item From the model:
\begin{itemize}
\item Disentangling two main mechanisms:
\begin{itemize}
\item Shotgun marriage couples are different in characteristics
\item Unexpected pregnancy creates large unplanned costs
\end{itemize}
\item Expensive eraly-life childcare explains a lot
\item Higher marginal costs of childcare for skilled women
\item $\sim10\%$ of couples marry because of unplanned pregnancy
\item These couples are the least stable
\end{itemize}
\item Missing pieces (TBD):
\begin{itemize}
\item Quantifying implications
\item Quantifying changes over time
\end{itemize}
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{Implications}
\begin{itemize}
\item Unintended pregnancies $\Rightarrow$ people to pick less favorable partners, ``inefficient'' marriages from ex-ante prospective
\item Child development:
\begin{itemize}
\item Single mothers or blended families $\Rightarrow$ inferior child's outcomes
\item Unstable marriages may be even worse but are often overlooked
\end{itemize}
\item Major impact on women, in particular on:
\begin{itemize}
\item Labor market choices and outcomes
\item Divorces + children worsen re-marriage prospective
\item Total fertility and marriage rates
\end{itemize}
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{Literature}
Shotgun marriages are still a thing:
\begin{itemize}
\item Gibson-Davis et al (2016): 10\% midpregnancy marriages
%\item Akerlof et al (1996)
\end{itemize}
Lifecycle models:
\begin{enumerate}
\item Endogenous bargaining, marriage, divorce \& assets:
\begin{itemize}
\item Voena (2015) --- just divorce, Low et al (2018) --- both
\item Mazzocco (2007) --- seminal paper
\end{itemize}
%\begin{center}
%\textbf{Contribution:} endogenous + exogenous fertility
%\end{center}
\item Fertility:
\begin{itemize}
\item Planned: Sommer (2016)
\item Unplanned: Ejrnæs, Jørgensen (2018)
\end{itemize}
%\begin{center}
%\textbf{Contribution:} bargaining, marriage and divorce
%\end{center}
\item Child investment and marriage/divorce:
\begin{itemize}
\item Lafortune, Low (2018)
\end{itemize}
%\begin{center}
%\textbf{Contribution:} proper lifecycle, fertility
%\end{center}
\end{enumerate}

\textit{Other relevant evidence:}
\begin{itemize}
\item Divorce rates by income: Boertien, Härkönen (2014)
\item Child expenditures by income: Schneider et al (2014)
\end{itemize}
%\item Computations: Judd et al (2014) (thank Luigi Bocola)

@@

\section{Model Features}

\begin{frame}
\frametitle{Overview}
\begin{itemize}
\item Lifecycle model
\item Assets ($a$) and productivity shocks ($z^f$, $z^m$)
\item Endogenous marriage and divorce (limited commitment)
\item Reduced form marriage market:
\[p^{\text{meet}}(\text{Age}) \text{ --- meeting a random partner with similar $a$ and $z$}\]
\item Both endogenous and unplanned fertility:
\[p^{\text{unplanned}}(\text{Age},z^f) \text{ --- arrival of unplanned kids}\]
\end{itemize}
\end{frame}

   \node [block, below = 1.5 cm of 2] (5) {\textbf{Kids First} \\[-0.5ex] \textbf{Couple}};
   \node [block, below = 0.8 cm of 3] (7) {No Kids \\[-0.5ex] Couple};
   \node [block, below = 0.8 cm of 7] (8) {\textbf{Mariage First} \\[-0.5ex] \textbf{Couple}};
   \draw[->] (1) to [bend left = 20] node [left] {\scriptsize Unplanned \\[-0.75ex] \scriptsize pregnancy} (2);
   \draw[->] (1) to [bend right = 20] node [right] {\scriptsize No \\[-0.75ex] \scriptsize pregnancy} (3);
   \draw[->] (2) to node [left] {\scriptsize Agree} node [right] {\scriptsize \textbf{Shotgun} \\[-0.75ex]  \scriptsize \textbf{marriage}} (5);
   \draw[->] (3) to node [right] {\scriptsize Agree} node [left]  {\scriptsize \textbf{Regular} \\[-0.75ex]  \scriptsize \textbf{marriage}} (7);
   \draw[->] (7) to node [right] {\scriptsize Kids by choice \\[-0.75ex] \scriptsize or randomly} (8);

    \node [block, above left   = 1.0 cm and -0.75 cm of 2] (4) {\,\, Single male\,\,};
    \node [block, above left   = 1.6 cm and -0.75 cm of 2] (4-sm) {Single mother};
	\node [block, above right   = 1.0 cm and -0.75 cm of 3] (6) {\,\, Single male\,\,};
    \node [block, above right   = 1.6 cm and -0.75 cm of 3] (6-f) {\,Single female\,};   
    \draw[->] (2.180) to [bend left = 30] node [left] {\scriptsize Disagree} (4.270);
    \draw[->] (3.0) to [bend right = 30] node [right] {\scriptsize Disagree} (6.270);
     \draw[->] (6.180) to [bend left = 30] node [above] {\scriptsize Back}  (1.-12);
   \draw[->] (6-f.180) to [bend right = 30] node [above] {\scriptsize Back} (1.12);
   \draw[->,thick,dashed] (4-sm.90) to [bend left = 30] node [above] {\scriptsize Back with delay}  (1.168);
   \draw[->] (4.0) to [bend right = 30] node [above] {\scriptsize Back} (1.192);

\begin{frame}[label=utilities]
\frametitle{Utility Functions}
\begin{itemize}
\item Couple, $\theta$ --- Pareto weight, $\psi$ --- common love shock:
\[U = \theta \cdot u(c_f) + (1-\theta)\cdot u(c_m) + \psi,\]
\item Couple + child, $Q$ --- child quality (public good):
\[U = \theta \cdot \tilde{u}(c_f,Q) + (1-\theta)\cdot \tilde{u}(c_m,Q) + \psi,\]
\item Child quality: $Q = F(x,l^f)$, $x$ --- money, $l^f$ --- female time
\item Budget constraint ($C$ reflects returns to scale, $a$ --- assets):
 \[C(c_f,c_m) + a' + x = R\cdot a + (1-l^f)\cdot W^f_t(z_f) + W^m_t(z_m),\]
\end{itemize}
\hyperlink{value-functions}{\beamerbutton{Value Functions}}
\end{frame}

\begin{frame}
\frametitle{Marriage}
\begin{itemize}
\item People (f \& m) meet, get a draw of love shock $\psi$ and agree to marry if exists a Pareto weight $\theta$ such that
\[V^{f,\text{yes}}(\underset{\scriptsize +}{\theta\vphantom{\psi}},\underset{\scriptsize +}{\psi}) \geq V^{f,\text{no}}, \ \ \ \ V^{m,\text{yes}}(\underset{\scriptsize  -}{\theta\vphantom{\psi}},\underset{\scriptsize +}{\psi}) \geq V^{m,\text{no}}.\]
\item Young children require lots of female time
\item \textbf{Mechanism:} unplanned pregnancy 
\begin{enumerate}
\item[(1)] lowers current $V^{f,\text{no}}$: hard to be a single mother
\item[(2)] increases future $V^{m,\text{yes}}$ and $V^{f,\text{yes}}$: good to have children
\end{enumerate} 
$(1) + (2)$ $\Rightarrow$ the couple agrees to marry with lower $\psi$ and $\theta$.
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{Divorce}
\begin{itemize}
\item Shocks to $\psi$, $z^f$ and $z^m$
\item People cannot commit not to divorce
\item $\theta$ may be renegotiated, stays constant if not
\item Divorce happens if no $\theta$ exist such that
\[V^{f,\text{together}}(\theta,\psi) \geq V^{f,\text{divorce}}, \ \ \ \ V^{m,\text{togetehr}}(\theta,\psi) \geq V^{m,\text{divorce}}.\]
\item Couples with lower $\psi$ or more unequal $z$ divorce more!
\item As children get older they require less time
\end{itemize}
\end{frame}

\section{Fit and Results}
\begin{frame}
\frametitle{Strategy}
\begin{itemize}
\item I use Simulated Method of Moments to calibrate $\sim 16$ parameters:
\begin{itemize}
\item Preferences
\item Probabilities of transitions
\item Child quality technology
\end{itemize}
\item Income paths to match ACS
\item The target moments are taken from:
\begin{itemize}
\item ACS (raw divorce, fertility and marriage shares by groups)
\item CEX + ATUS (time and money spent on childcare)
\end{itemize}
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{Lower Child Investment Even without Selection2}
\vspace{-2cm}
\begin{center}
\includegraphics[width=1.0\linewidth]{../main/x-time.pdf}
\end{center}
\end{frame}

\begin{frame}
\frametitle{Conclusions And Plans}
\begin{itemize}
\item What we observe about childcare costs generates the pattern
\item KF couples have lower $\psi$ and divorce more
\item Unplanned pregnancies account for $1/2$ of early-age divorces
\item Preparing a new iteration now:
\begin{itemize}
\item TBD: partition by income needs to be improved
\item TBD: decompose by different contributors ($z$, $\psi$, age)
\end{itemize}
\item Model allows to account for other stories:
\begin{itemize}
\item Labor market discrimination of unmarried mothers
\item Social stigma against unmarried cohabitation
\item ...
\end{itemize}
\end{itemize}

@@
